---
title: "EDA"
author: "Zihan Zhou"
date: "2025-11-25"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: cosmo
    css: styles.css
---

<a href="index_citibike.html" class="back-home">
  <i class="fa fa-home"></i> Back to Home
</a>

<div class="hero-section">
  <h1>📊 Exploratory Data Analysis</h1>
  <p>Uncovering Patterns in NYC Citi Bike Usage</p>
</div>

<br>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<div class="info-card">

# Data Loading
Load key packages and data. 

```{r}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(gridExtra)
library(tsibble)
library(feasts)
library(patchwork)
library(scales)
library(knitr)
library(broom)
library(forecast)

bike <- read_csv("data/final_bike_weather_categorized.csv")

glimpse(bike)
```

</div>

<div class="info-card">

# 1. Dataset Overview

The dataset contains **stratified sampled Citi Bike trips from September 2025**, enriched with **weather categories**.  
Each row represents **a single bike ride** with detailed trip-level information including:

- Ride ID  
- Bike type (electric vs. classic)  
- Start/end time and station  
- Rider type (member vs. casual)  
- Ride duration (hours), distance (km), bearing  
- Weather on the ride date (tmax, tmin, prcp, temperature category, rain category)  
- Weekend/weekday indicator  

This allows us to analyze:

- Time-of-day and weekday/weekend behavior  
- Member vs. casual behavior  
- Weather effects  
- Time-series trends across the month  

In this part, we perform exploratory data analysis (EDA) and visualization for Citi Bike trips in September 2025 focusing on temporal patterns and weather influences. 

```{r}
if ("datetime_start_obj" %in% names(bike) && !all(is.na(bike$datetime_start_obj))) {
  bike <- bike %>%
  mutate(datetime_start = ymd_hms(datetime_start_obj, quiet = TRUE))
} else {
  bike <- bike %>% mutate(datetime_start = ymd_hms(started_at, quiet = TRUE))
}

bike <- bike %>%
  mutate(
  date = as.Date(datetime_start),
  hour = hour(datetime_start),
  wday = wday(datetime_start, label = TRUE, week_start = 1),
  weekday_weekend = case_when(
  weekend_and_holiday %in% c("weekend", "weekend and holiday", "Weekend", "weekend and holiday") ~ "Weekend", TRUE ~ "Weekday"),
  user_type = member_casual
)
```

</div>

<div class="info-card">

# 2. Daily Aggregates

```{r}
daily <- bike %>%
  group_by(start_date) %>%
  summarise(
    total_rides = n(),
    avg_duration = mean(ride_duration_hours, na.rm = TRUE),
    member_rides = sum(member_casual == "member"),
    casual_rides = sum(member_casual == "casual"),
    tmax = tmax,
    prcp = prcp
  )

head(daily)
```

</div>

<div class="info-card">

# 3. Daily Ride Volume

```{r}
p_daily_rides <- ggplot(daily, aes(x = start_date, y = total_rides)) +
  geom_line(size = 1) +
  labs(title = "Daily Ride Volume — September 2025",
       x = "Date", y = "Ride Count") +
  theme_minimal()

p_daily_rides
```

</div>

<div class="info-card">

# 4. Member vs Casual Ride Patterns

## 4.1 Daily Average Duration

```{r}
dur_ts <- bike %>%
  group_by(start_date, member_casual) %>%
  summarise(mean_duration = mean(ride_duration_hours, na.rm = TRUE))

ggplot(dur_ts, aes(start_date, mean_duration, color = member_casual)) +
  geom_line(size = 1) +
  labs(title = "Daily Average Ride Duration — Member vs Casual",
       x = "Date", y = "Duration (hours)") +
  theme_minimal()
```

---

## 4.2 Hourly Ride Count

```{r}
hourly <- bike %>%
  group_by(hour, member_casual, weekend_and_holiday) %>%
  summarise(rides = n())

ggplot(hourly, aes(hour, rides, color = member_casual)) +
  geom_line(size = 1) +
  facet_wrap(~ weekend_and_holiday) +
  labs(title = "Hourly Ride Patterns — Member vs Casual (Weekday vs Weekend)",
       x = "Hour of Day", y = "Ride Count") +
  theme_minimal()
```

---

## 4.3 Day of Week Pattern

```{r}
wday_daily <- bike %>%
  group_by(wday, user_type) %>%
  summarise(rides = n(), .groups = "drop") %>%
  mutate(wday = factor(wday, levels = c("Mon","Tue","Wed","Thu","Fri","Sat","Sun")))

p_wday <- ggplot(wday_daily, aes(x = wday, y = rides, fill = user_type)) +
  geom_col(position = "dodge") +
  labs(title = "Ridership by Day of Week (member vs casual)",
       x = "Day of Week", y = "Ride Count", fill = "User Type") +
  theme_minimal()

p_wday

```

As shown in the bar chart, it is clear that member users have much higher ride counts than casual users.

Specifically, member ride counts are particularly high on Mondays and Tuesdays and reach their lowest on weekends. 

In contrast, casual users' ride counts are relatively consistent throughout the week, with slightly higher numbers on weekends. The observation confirms the previous conclusion.


```{r}
p_dur_group <- bike %>%
  group_by(member_casual, weekend_and_holiday) %>%
  summarise(
    mean_duration = mean(ride_duration_hours, na.rm = TRUE)
  ) %>%
  ggplot(aes(x = weekend_and_holiday, y = mean_duration,
             fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(title = "Average Ride Duration — Member vs Casual across Weekday/Weekend",
       x = "Day Type", y = "Duration (hours)") +
  theme_minimal()

p_dur_group
```

</div>

<div class="info-card">

# 5. Time Series of Average Ride Duration

```{r}
dur_ts <- bike %>%
  group_by(start_date, member_casual) %>%
  summarise(mean_duration = mean(ride_duration_hours, na.rm = TRUE))

p_dur_ts <- ggplot(dur_ts, aes(start_date, mean_duration, color = member_casual)) +
  geom_line(size = 1) +
  labs(title = "Daily Average Ride Duration — Member vs Casual",
       x = "Date", y = "Duration (hours)") +
  theme_minimal()

p_dur_ts
```

</div>

<div class="info-card">

# 6. Weather Effects

## 6.1 Ride Count vs Temperature Level Category

```{r}
p_temp_level <- bike %>%
  group_by(start_date, temp_level_category) %>%
  summarise(rides = n()) %>%
  ggplot(aes(temp_level_category, rides)) +
  geom_boxplot() +
  labs(title = "Ride Counts by Temperature Level Category",
       x = "Temperature Category", y = "Daily Ride Count") +
  theme_minimal()

p_temp_level
```

This plot shows that higher temperatures are associated with increased ridership.

---

## 6.2 Ride Count vs Rain Category

```{r}
p_rain <- bike %>%
  group_by(start_date, rain_category) %>%
  summarise(rides = n()) %>%
  ggplot(aes(rain_category, rides)) +
  geom_boxplot() +
  labs(title = "Ride Counts by Rain Category",
       x = "Rain Category", y = "Daily Ride Count") +
  theme_minimal()

p_rain
```

The analysis reveals a negative correlation between rainfall and ridership.

---

## 6.3 Weather Variables

```{r}
# tmax
ggplot(daily, aes(x = tmax, y = total_rides)) +
  geom_point(alpha = .4) +
  geom_smooth(method = "loess") +
  labs(title = "Ride Count vs Daily Maximum Temperature",
       x = "tmax", y = "Rides") +
  theme_minimal()

# prcp
ggplot(daily, aes(x = prcp, y = total_rides)) +
  geom_point(alpha = .4) +
  geom_smooth(method = "loess") +
  labs(title = "Ride Count vs Precipitation",
       x = "Precipitation (mm)", y = "Rides") +
  theme_minimal()
```

</div>

<div class="info-card">

# 7. Bike Type Analysis

```{r}
p_type <- bike %>%
  group_by(start_date, rideable_type) %>%
  summarise(rides = n()) %>%
  ggplot(aes(rideable_type, rides)) +
  geom_boxplot() +
  labs(title = "Ride Duration by Bike Type",
       x = "Bike Type", y = "Ride Counts") +
  theme_minimal()

p_type
```


</div>

<div class="info-card">

# 8. Time Series Analysis

```{r}
library(tseries)
daily_rides <- bike %>%
  group_by(start_date) %>%
  summarise(total_rides = n(), .groups = "drop")

library(tsibble)
library(fable)

ts_data <- daily_rides %>%
  as_tsibble(index = start_date)

autoplot(ts_data, total_rides) +
  labs(title = "Daily Ride Counts",
       x = "Date",
       y = "Number of Rides")

```

```{r}
fit_arima <- ts_data %>%
  model(ARIMA(total_rides))

report(fit_arima)

```

```{r}
fc <- fit_arima %>% forecast(h = 7)

autoplot(fc, ts_data) +
  labs(title = "ARIMA Forecast of Daily Rides",
       x = "Date",
       y = "Number of Rides")

```

---

## 8.1 Decomposition

```{r}
fit_ets <- ts_data %>% model(ETS(total_rides))

components_plot <- autoplot(components(fit_ets)) +
  labs(
    title = "ETS Decomposition of Daily Ride Counts",
    subtitle = "Using Exponential Smoothing (suitable for short, non-periodic series)",
    x = "Date",
    y = "Value"
  ) +
  theme_minimal()

components_plot

```

</div>

<div class="highlight-box">

# Summary

- Member and casual riders display distinct **daily and weekend/weekday patterns**.  
- Daily ride volume shows clear **temporal trends** across September.  
- Temperature and rainfall strongly shape both **trip volume**.  
- Time-series decomposition reveals trend + seasonality suitable for forecasting. 

</div>

<br>

<a href="index_citibike.html" class="back-home">
  <i class="fa fa-home"></i> Back to Home
</a>



